MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments

### 2022-03-05 main/MSstatsPTM_Main.tex
experimental design changes

The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are facilitated by the modification of proteins at specific residues, acting as molecular on/off switches\cite{Deribe,Cohen}. Bottom-up liquid chromatography coupled with mass spectrometry (LC-MS) is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) \cite{Kall:2011ub,Roepstorff}. Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site's occupancy repertoire, or relative changes in occupancy across experimental conditions\cite{Mann}. Some scientific questions,  such as discovery of protein biomarkers of a disease, consider meaningful patterns of changes in PTM abundance alone \cite{Petushkova_2017}. Other questions, e.g. in biological and physiological studies based on very short timescales, require us to distinguish the confounding of quantitative changes in PTM and of overall changes of the protein abundance.

@@

Currently statistical analyses of these experiments are performed with methods such as t-test and Limma\cite{Ritchie_15a}. These approaches generally use the intensity ratios of modified and unmodified peptide features and compare the mean abundance of different PTM sites. These approaches do not fully account for different sources of variations and as such, are not directly applicable to experiments with complex designs, such as multiple conditions and mixtures. Additionally, there is no self contained implementation of these methods, requiring manual application of the methods by the user and increasing complexity.

This manuscript proposes a general statistical analysis framework that detects relative changes in post-translational modifications. The proposed framework combines the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, and takes into account various sources of variation and confounding. The framework requires an experimental design that, for each sample, quantifies the peptides with post-translational modifications, as well as  peptides from the same proteins with no modification sites. It is applicable to both label-free and tandem mass tag (TMT)-based acquisitions. The proposed framework summarizes the abundances of peptides with and without the modification sites, and fits a linear mixed effects model that automatically adjusts to the biological and technological aspects of the experimental design. Finally changes in the PTM and protein-level abundances are combined to remove bias due to the confounding of these two sources.

@@

The proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT\cite{Choi:2014,Huang:2020}.

@@

This section summarizes the datasets in this manuscript. The datasets were chosen to provide a performance benchmark of the proposed approach in situations with known ground truth, and to represent a variety of experimental designs and acquisition methods. Two computer simulations were run which varied in experimental realism. The first simulation was perfectly clean, with many replicates and no missing values, while the second simulation introduced real world qualities, such as limited modified features and missing values. The spike-in experiment took the real world qualities a step further and allowed us to compare the methods in a real experiment with known changes in modified spike-in peptides. Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, experimental designs, and data acquisition strategies. A summary of the experiments used in this manuscript can be seen in Table \ref{tab:dataDescription}. Details of raw data, R scripts with $MSstatsPTM$ analyses, and result of statistical analyses for the datasets are available in either MassIVE.quant or Github (with direct links available in the table).

%%%% Summary of controlled mixtures
\begin{table}[h!]
\centering
\begin{tiny}
\begin{tabular}{|c|c|cccc|c|}
\hline
%%%% colnames of table
&  & \multicolumn{4}{c|}{Experimental design} & Data availability \\
& Dataset & \multirow{2}{*}{Comparison} & No. of & No. of Bio.& \multirow{2}{*}{Details in} & (MassIVE.quant  \\
&  & & Conditions & Replicates &  & or R package) \\
\hline
\hline
 &&&&&& \\[-0.05in]
%%%% DDA:ControlledMix
\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group & 2/3/4 & 2/3/5/10 & Suppl. Sec. \ref{sec:dataset1} & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}\\
 &&&&&& \\%[-0.05in]
%\hline 
%&&&&&&&& \\[-0.05in]
% \cline{2-9}
% &&&&&&&& \\[-0.05in]
%%% DDA iPRG2015
& Computer simulation 2 - missing and low features & Group & 2/3/4 & 2/3/5/10 & Suppl. Sec. \ref{sec:dataset2} & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
 &&&&&& \\%[0.05in]
%\hline 
% &&&&&& \\[-0.05in]
%%% DDA Cox2014
& SpikeIn benchmark & Group & 4 & 2 & Suppl. Sec. \ref{sec:benchmark} & \\
 &&&&&& \\%[-0.05in]
\hline
\multicolumn{7}{c}{ } \\ [0.02in]
\hline 
 &&&&&& \\%[-0.05in]
%%%%%%%%%%%%%%%%%%
%%%% biological study
%%%%%%%%%%%%%%%%%%
%% DDA Meierhofer
\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Investigation}} & Human - Ubiquitination - TMT & Group & 6 & 2 & Suppl. Sec. \ref{sec:ipah} &  \\
&&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Picotti 2009
& Mouse - Phosphorylation - TMT & Time course & 6 & 2 & Suppl. Sec. \ref{sec:shigella} & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} \\
&&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Surinova 2015, training
& Human - Ubiquitination - Label-free & Group & 4 & 2 & Suppl. Sec. \ref{sec:usp30} &\\ [0.02in]
 &&&&&& \\
 \hline
\end{tabular}\\
\end{tiny}
\caption{ \small {\bf Experimental datasets in this manuscript} 
```Dataset'' is the code name of the dataset used in this manuscript.
`Comparison'' is the type of experimental design (group comparison, or time course).
``No. of Conditions'' states the number of conditions in the dataset.
``No. of Bio. Replicates'' states the number of biological replicates per condition.
Details regarding the experimental design for each dataset are available in the section listed in the ``Details in'' column.
``Data availability'' column shows the MassIVE.quant ID or github repository. 
}
\label{tab:dataDescription}
\end{table}

@@

To evaluate the proposed method in situations with biological variation and known ground truth we created synthetic datasets which mimicked a label-free experiment; with a high number of modified peptide features and no missing values. Multiple simulations were generated with differing numbers of replicates and conditions per simulation. This was done to see how the method performed in different experimental design scenarios. The exact numbers of replicates and conditions can be seen in Table \ref{tab:dataDescription}.

In each simulation, 1000 PTMs and 1000 unmodified proteins were generated. Each PTM and protein was created with 10 corresponding peptide features. 500 of the PTMs were generated with a differential fold change between conditions, while the other half were generated with no difference between conditions. Of the 500 differential PTMs, the fold change of 250 were fully masked by changes in the unmodified protein. Of the 500 non-differential PTMs, 250 were generated with a differential fold change that was entirely due to the unmodified protein. All differential PTMs were generated with an expected log fold change of .75. Each simulation was generated with random biological variation. Specifically, let $Y_{ibc}$ denote the abundance of peptide $i$ in biological replicate $b$ of condition $c$. The simulated peptide abundance  $Z_{ibc}$ was generated as $Z_{ibc} = Y_{ibc} + \epsilon_{ibc}$ where $\epsilon_{ibc}\sim^{iid}N(0,\sigma^2)$. The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$ motivated by the biological experiments in this manuscript.

@@

In this simulation, data was generated using the same parameters as described in Section \ref{sec:comp_sim_procedure1} and including real world experimental conditions. Specifically these simulations included missing values and a low number of modified features. The percentage of missing values and feature counts were determined by looking at the biological experiments in this paper. PTMs were simulated with 2 modified peptide features, while unmodified proteins were still simulated with 10 features. Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random. Adding missing values and few representative modified features provides a more realistic expectation of model performance.

@@

\subsection*{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free}

This custom designed experiment had 4 conditions with two biological replicates per condition and a balanced design.

Experimental Design: We evaluated our approach using a custom designed spike-in benchmark experiment with known ground truth, where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance of these 20 proteins were the target of the benchmark.  Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes. Background E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling. The background lysate was treated as controls and are not expected to be differential in any comparison. The spike-in peptides were mixed with human lysate to create four mixture conditions. Two sets of data were acquired for each mixture: KGG enriched + LC-MS, and LC-MS only. The four mixtures were compared with known fold changes for the spike-in peptides.

More information on the experimental design of the benchmark experiment can be found in Supplementary Sec. \ref{supp-sec:benchmark}.

\subsection*{Dataset 4 : Human - Ubiquitination - 1mix-TMT}

The proposed approach was evaluated on an experiment where Shigella ubiquitin ligase IpaH7.8 was shown to degrade the protein gasdermin D (GSDMD)\cite{LUCHETTI2021}. Multiplex proteomics was used to quantify the abundance of total protein and ubiquitination in human epithelial cells. Cells were engineered to express IpaH7.8 under a dox inducible promoter and measurements were taken at different time periods. GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. The experiment was unbalanced with two bioreplicates per condition for all conditions except for infected 1 hour. About ~95\% of the identified modified peptides derived from proteins that were quantified in the global profiling run.

@@

\subsection*{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}

Here an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri)\cite{Maculins}. Multiplex proteomics was used to quantify the abundance of total protein, and phosphorylation in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri. The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours). Quantifying the total protein along with the post-translational modifications allowed us to adjust for changes in total protein and see the true impact of the site specific modifications. Two mixtures using 11-plex were ran over the six conditions. The six conditions were split between 11 channels leading to the experimental design being unbalanced. Each mixture contained two replicates per early and late WT and KO conditions. Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO. Mixture two contained one replicate of uninfected KO and two uninfected WT. About ~90\% of the identified modified peptides derived from proteins that were also quantified in the global profiling run.

@@

\subsection*{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run}

This experiment looked into the relationship between USP30 and protein kinase PINK1, and their association with Parkinson’s Disease. Ubiquitination site profiling was performed and the modified site abundance was analyzed. Four conditions were tested with two biological replicates per condition. The conditions were as follows: CCCP, USP30 over expression (USP30 OE), Combo, and Control. Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides. In contrast to the other biological experiments assessed in this paper, this experiment did not not have a separate global profiling run for the unmodified protein. In this case $MSstatsPTM$ can still be used by extracting unmodified peptides from the modified run, however this generally leads to substantially less matches between modified and unmodified peptides, in addition to low feature counts for unmodified peptides. In this case there only 41.9\% of the modified peptides had a corresponding unmodified protein to perform adjustment. In cases where there is no estimate for the matching unmodified protein, adjustment cannot be performed.

@@

Despite the important implications of PTMs in biological functions, there is a lack of general framework to summarize the available quantitative information from LC-MS data, to perform statistical inference, and to draw conclusions to characterize the quantitative properties of PTM in a statistically rigorous manner. Many investigations performed differential abundance analysis of PTMs using two-sample t-test or its extensions. The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another. Modifications of the t-test such as moderated t-test with Limma were also proposed\cite{Zhu}. While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc. Additionally, while this approach can be applied to experiments targeting PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.

@@

The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}. It takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance\cite{Choi:2014}. This approach allows the method to be extended to cases where the experimental design is unbalanced, and where additional sources of variation are present. In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site\cite{Tukey:1977}. After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models\cite{Bolker2009} \cite{Faraway:2006}. Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.

Additionally, the proposed method can include imputing missing feature intensities if desired. When values are imputed, it is assumed they are missing for reasons of low abundance. Missing features are imputed in each MS run using the Accelerated Time Failure (AFT) model\cite{Tukey:1977}. In order to impute a feature's missing values, the feature must be present in at least one MS Run. If the feature is not present across all MS Runs, the value will be left missing. Missing value imputation is done before summarization with Tukey's Median Polish in order to correct for the impact of outliers. Missing value imputation is done separately for both the PTM and global protein datasets.

@@

The statistical modeling approaches discussed above can also be extended to complex designs including experiments with additional sources of variation and unbalanced designs. Equation (1) is the same in this case and the null hypothesis is still that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$.

Model inference is done in the same way to how MSstats~\cite{Choi:2014} and MSstatsTMT~\cite{Huang:2020} target the global protein. In cases where the design is unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood (RML) is used to estimate the parameters of each model. Once the parameters are estimated, we can combine the models using modified versions of the formulas in Section \ref{sec:adjustment_summary}.

The log-fold change in the adjusted PTM abundance, $\Delta$, is now estimated by RML, and Equation (2) becomes 
\begin{equation}
\hat{\Delta} = (\hat{\mu}_{RML_i}^{\ast} - \hat{\mu}_{RML_{i'}}^{\ast}) - (\hat{\mu}_{RML_i} - \hat{\mu}_{RML_{i'}})
\end{equation}

The standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ and degrees of freedom in Equations (3) and (4) are unchanged from the balanced case, although $\hat{\sigma}_{\gamma^z\ast}$ and $\hat{\sigma}_{\gamma}$ are now estimated using RML.

@@

\subsubsection*{In label-free experiment without a separate global profiling run, the proposed approach could eliminate the bias \todo{due to what?}, albeit less effectively than \todo{in presence of a global profiling run?}}

>>>>

The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are facilitated by the modification of proteins at specific residues, acting as molecular on/off switches \cite{Deribe,Cohen}. Bottom-up liquid chromatography coupled with mass spectrometry (LC-MS) is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) \cite{Kall:2011ub,Roepstorff}. Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site's occupancy repertoire, or relative changes in occupancy across experimental conditions \cite{Mann}. Some scientific questions,  such as discovery of protein biomarkers of a disease, consider meaningful patterns of changes in PTM abundance alone \cite{Petushkova_2017}. Other questions, e.g. in biological and physiological studies based on very short timescales, require us to distinguish the confounding of quantitative changes in PTM and of overall changes of the protein abundance.

@@

Currently statistical analyses of these experiments are performed with methods such as t-test and Limma \cite{Ritchie_15a}. These approaches generally use the intensity ratios of modified and unmodified peptide features and compare the mean abundance of different PTM sites. These approaches do not fully account for different sources of variations and as such, are not directly applicable to experiments with complex designs, such as multiple conditions and mixtures. Additionally, there is no self contained implementation of these methods, requiring manual application of the methods by the user and increasing complexity.

This manuscript proposes a general statistical analysis framework that detects relative changes in post-translational modifications. The proposed framework combines the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, and takes into account various sources of variation and confounding. The framework requires an experimental design that, for each sample, quantifies the peptides with post-translational modifications, as well as  peptides from the same proteins with no modification sites. It is applicable to both label-free and tandem mass tag (TMT)-based acquisitions. The proposed framework summarizes the abundances of peptides with and without the modification sites, and fits a linear mixed effects model that automatically adjusts to the biological and technological aspects of the experimental design. Finally, changes in the PTM and protein-level abundances are combined to remove bias due to the confounding of these two sources.

@@

The proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT \cite{Choi:2014,Huang:2020}.

@@

This section summarizes the datasets in this manuscript. The datasets were chosen to provide a performance benchmark of the proposed approach in situations with known ground truth, and to represent a variety of experimental designs and acquisition methods. Two computer simulations were run which varied in experimental realism. The first simulation was perfectly clean, with many replicates and no missing values, while the second simulation introduced real world qualities, such as limited modified features and missing values. The spike-in experiment took the real world qualities a step further and allowed us to compare the methods in a real experiment with known changes in modified spike-in peptides. Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, experimental designs, and data acquisition strategies. A summary of the experiments used in this manuscript can be seen in Table \ref{tab:dataDescription}. Details of the raw data, R scripts with $MSstatsPTM$ analysis, and the results of the statistical analysis are available in either MassIVE.quant or Github (with direct links available in the table).

%%%% Summary of controlled mixtures
\begin{table}[h!]
\centering
\begin{tiny}
\begin{tabular}{|c|c|ccc|c|}
\hline
%%%% colnames of table
&  & \multicolumn{3}{c|}{Experimental design} & Data availability \\
& Dataset & \multirow{2}{*}{Comparison} & No. of & No. of Bio. & (MassIVE.quant  \\
&  & & Conditions & Replicates & or R package) \\
\hline
\hline
 &&&&& \\[-0.05in]
%%%% DDA:ControlledMix
\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}\\
 &&&&& \\%[-0.05in]
%\hline 
%&&&&&&&& \\[-0.05in]
% \cline{2-9}
% &&&&&&&& \\[-0.05in]
%%% DDA iPRG2015
& Computer simulation 2 - missing and low features & Group & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
 &&&&& \\%[0.05in]
%\hline 
% &&&&&& \\[-0.05in]
%%% DDA Cox2014
& SpikeIn benchmark - Ubiquitination - Label-free & Group & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971}\\
 &&&&& \\%[-0.05in]
\hline
\multicolumn{6lm}{c}{ } \\ [0.02in]
\hline 
 &&&&& \\%[-0.05in]
%%%%%%%%%%%%%%%%%%
%%%% biological study
%%%%%%%%%%%%%%%%%%
%% DDA Meierhofer
\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Investigation}} & Human - Ubiquitination - 1mix-TMT & Group & 6 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Picotti 2009
& Mouse - Phosphorylation - 2mix-TMT & Time course & 6 & 4 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Surinova 2015, training
& Human - Ubiquitination - Label-free & Group & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977}\\ [0.02in]
 &&&&& \\
 \hline
\end{tabular}\\
\end{tiny}
\caption{ \small {\bf Experimental datasets in this manuscript} 
```Dataset'' is the code name of the dataset used in this manuscript.
`Comparison'' is the type of experimental design (group comparison, or time course).
``No. of Conditions'' states the number of conditions in the dataset.
``No. of Bio. Replicates'' states the number of biological replicates per condition.
``Data availability'' column shows the MassIVE.quant ID or github repository. 
}
\label{tab:dataDescription}
\end{table}

@@

To evaluate the proposed method in situations with biological variation and known ground truth we created synthetic datasets which mimicked a label-free experiment. Multiple simulations were generated with differing numbers of replicates and conditions per simulation. All simulations were created with a high number of modified peptide features and no missing values. This was done to see how the method performed in different experimental design scenarios. The exact numbers of replicates and conditions can be seen in Table \ref{tab:dataDescription}.

In each simulation, 1000 PTMs and 1000 unmodified proteins were generated. Each PTM and protein was created with 10 corresponding features. 500 of the PTMs were generated with a differential fold change between conditions, while the other half were generated with no difference between conditions. Of the 500 differential PTMs, the fold change of 250 were fully masked by changes in the unmodified protein. Of the 500 non-differential PTMs, 250 were generated with a differential fold change that was entirely due to the unmodified protein. All differential PTMs were generated with an expected log fold change of .75 between conditions. Each simulation was generated with random biological variation. Specifically, let $Y_{ibc}$ denote the abundance of peptide $i$ in biological replicate $b$ of condition $c$. The simulated peptide abundance  $Z_{ibc}$ was generated as $Z_{ibc} = Y_{ibc} + \epsilon_{ibc}$ where $\epsilon_{ibc}\sim^{iid}N(0,\sigma^2)$. The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$ motivated by the biological experiments in this manuscript.

@@

In this simulation, data was generated using the same parameters as described in the previous section and including real world experimental conditions. Specifically these simulations included missing values and a low number of modified features. The percentage of missing values and feature counts were determined by looking at the biological experiments in this paper. PTMs were simulated with 2 modified peptide features, while unmodified proteins were still simulated with 10 features. Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random. Adding missing values and few representative modified features provides a more realistic expectation of model performance.

@@

\subsection*{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free}
\label{sec:exp_proc_dataset3}
We evaluated our approach using a custom designed spike-in benchmark experiment with known ground truth. This experiment had 4 conditions with two biological replicates per condition and a balanced design.

Experimental Design: In this experiment 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance of these 20 proteins were the target of the benchmark.  Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes. Background E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling. The background lysate was treated as controls and are not expected to be differential in any comparison. The spike-in peptides were mixed with human lysate to create four mixture conditions. Two sets of data were acquired for each mixture: KGG enriched + LC-MS, and LC-MS only.

Global Profiling Run: This experiment included a separate global profiling run to measure unmodified peptides. There was a 90.2\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.

Pairwise Comparisons: We evaluated the ability of the methods to detect known changes in abundance between pairs of conditions. The four mixtures were treated as individual conditions and labeled mix1, mix2, mix3, and mix4. The pairwise comparisons were mix1-mix2, mix1-mix3, mix1-mix4, mix2-mix3, mix2-mix4, and mix3-mix4. The true log fold changes of spike-in modified peptides in these comparisons can be seen in Table \ref{table:spikein_fold_change}.

\begin{table}[h!]
\centering
\begin{tabular}{| c | c |}
\hline
 Comparison & Expected log fold change\\ [0.5ex]
 \hline\hline
 mix1-mix2 & -1\\
 \hline
 mix1-mix3 & 1\\
\hline
 mix1-mix4 & 0\\
\hline
 mix2-mix3 & -2\\
\hline
 mix2-mix4 & -1\\
\hline
 mix3-mix4 & 1\\
\hline
\end{tabular}
\caption{The expected log fold change of modified spike-in peptides in Dataset 3.}
\label{table:spikein_fold_change}
\end{table}

More information on the experimental design of the benchmark experiment and how the expected fold change was calculated can be found in Supplementary Sec. \ref{supp-sec:benchmark}.

\subsection*{Dataset 4 : Human - Ubiquitination - 1mix-TMT}
\label{sec:exp_proc_dataset4}
The proposed approach was evaluated on an experiment with six conditions and an unbalanced design.

Experimental Design: In this study Luchetti et al. \cite{LUCHETTI2021} used multiplex proteomics to quantify the abundance of total protein and ubiquitination in human epithelial cells. Cells were engineered to express IpaH7.8 under a dox inducible promoter and measurements were taken at different time periods. GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.

Global Profiling Run: This experiment included a separate global profiling run to measure unmodified peptides. There was a 95\% overlap between the identified modified peptides and proteins that were quantified in the global profiling run.

Pairwise comparisons: We evaluated the statistical approach by its ability to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance. The six condition were labeled Dox1hr, Dox2hr, Dox4hr, Dox6hr, NoDox0hr, and NoDox6hr. All conditions were compared with each other in a full pairwise comparison, resulting in 15 comparisons. Because the dataset is a biological investigation, the true positive modifications were unknown.

@@

\subsection*{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}
\label{sec:exp_proc_dataset5}
In this experiment 8 biological replicates were allocated to 2 TMT mixtures in an unbalanced design.

Experimental Design: Maculins et al. \cite{Maculins} designed an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri). Multiplex proteomics was used to quantify the abundance of total protein, and phosphorylation in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri. The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours). 11-plex isobaric multiplexing via tandem mass tagging (TMT) in combination with liquid-chromatography and tandem mass spectrometry (LC-MS/MS) was used. Six conditions were split between 11 channels leading to the experimental design being unbalanced. Each mixture contained two replicates per early and late WT and KO conditions. Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO. Mixture two contained one replicate of uninfected KO and two uninfected WT.

Global Profiling Run: This experiment included a separate global profiling run to measure unmodified peptides. There was a 90\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.

Time series comparison: We evaluated the statistical approach by its ability to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance. The six condition were labeled KO\_Uninfect, KO\_Early, KO\_Late, WT\_Uninfect, WT\_Early, and WT\_Late. 9 total comparisons were made named KO\_Early-WT\_Early, KO\_Late-WT\_Late, KO\_Uninfected-WT\_Uninfected, KO\_Early-KO\_Uninfected, KO\_Late-KO\_Uninfected, WT\_Early-WT\_Uninfected, WT\_Late-WT\_Uninfected, Infected-Uninfected, and KO-WT. Because the dataset is a biological investigation, the true positive modifications were unknown.

@@

\subsection*{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run}
\label{sec:exp_proc_dataset6}

In this experiment 8 biological replicates were split between 4 conditions in a balanced design.

Experimental Design: Cunningham et al. \cite{Cunningham2015} investigated the relationship between USP30 and protein kinase PINK1, and their association with Parkinson’s Disease. Ubiquitination site profiling was performed and the modified site abundance was analyzed. Four conditions were tested with two biological replicates per condition. The conditions were as follows: CCCP, USP30 over expression (USP30 OE), Combo, and Control. Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides.

Global Profiling Run: This experiment did not included a separate global profiling run to measure unmodified peptides. In this case $MSstatsPTM$ can still be used by extracting unmodified peptides from the modified run, however this leads to substantially less matches between modified and unmodified peptides, in addition to low feature counts for unmodified peptides. There was a 41.9\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.

Pairwise comparisons: We evaluated the statistical approach by its ability to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance. The four condition were labeled CCCP, Combo, Ctrl, and USP30\_OE. All conditions were compared with each other in a full pairwise comparison, resulting in 6 comparisons. Because the dataset is a biological investigation, the true positive modifications were unknown.

@@

Despite the important implications of PTMs in biological functions, there is a lack of general framework to summarize the available quantitative information from LC-MS data, to perform statistical inference, and to draw conclusions to characterize the quantitative properties of PTM in a statistically rigorous manner. Many investigations performed differential abundance analysis of PTMs using two-sample t-test or its extensions. The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another. Modifications of the t-test such as moderated t-test with Limma were also proposed \cite{Zhu}. While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc. Additionally, while this approach can be applied to experiments targeting PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.

@@

The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}. It takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance \cite{Choi:2014}. This approach allows the method to be extended to cases where the experimental design is unbalanced, and where additional sources of variation are present. In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site \cite{Tukey:1977}. After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models \cite{Bolker2009} \cite{Faraway:2006}. Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.

Additionally, the proposed method can include imputing missing feature intensities if desired. When values are imputed, it is assumed they are missing for reasons of low abundance. Missing features are imputed in each MS run using the Accelerated Time Failure (AFT) model \cite{Tukey:1977}. In order to impute a feature's missing values, the feature must be present in at least one MS Run. If the feature is not present across all MS Runs, the value will be left missing. Missing value imputation is done before summarization with Tukey's Median Polish in order to correct for the impact of outliers. Missing value imputation is done separately for both the PTM and global protein datasets.

@@

The statistical modeling approaches discussed above can also be extended to complex designs including experiments with additional sources of variation and unbalanced designs. Equation 1 is the same in this case and the null hypothesis is still that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$.

Model inference is done in the same way to how MSstats~\cite{Choi:2014} and MSstatsTMT~\cite{Huang:2020} target the global protein. In cases where the design is unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood (RML) is used to estimate the parameters of each model. Once the parameters are estimated, we can combine the models using modified versions of Equations 2, 3, and 4.

The log-fold change in the adjusted PTM abundance, $\Delta$, is now estimated by RML, and Equation 2 becomes 
\begin{equation}
\hat{\Delta} = (\hat{\mu}_{RML_i}^{\ast} - \hat{\mu}_{RML_{i'}}^{\ast}) - (\hat{\mu}_{RML_i} - \hat{\mu}_{RML_{i'}})
\end{equation}

The standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ and degrees of freedom in Equations 3 and 4 are unchanged from the balanced case, although $\hat{\sigma}_{\gamma^z\ast}$ and $\hat{\sigma}_{\gamma}$ are now estimated using RML.

@@

\subsubsection*{In label-free experiment without a separate global profiling run, the proposed approach could eliminate the bias due to changes in the unmodified protein, albeit less effectively than in the presence of a global profiling run}
